---
title: "Bank of England Data Vis Competition - EDA"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Aim: Perform EDA on Bank of England data to come up with a story to visualise. 

```{r intro, echo=FALSE}

setwd('/Users/bryonymiles/documents/dataprojects/eda')
bank <-read.csv('bankdata.csv')

```

#1. Initial Analysis: structure of the dataset.

Dimensions?
```{r dim, echo=FALSE}
dim(bank)
```
Variables?
```{r vars, echo=FALSE}
names(bank)
```

Field Descriptions:

1. **Year**
2. **deficit**: Central government deficit (Calendar Year Totals, £mn)
3. **surplus**: Central government primary surplus (Calendar Year Totals, £mn)
4. **GDP**: Real GDP at market prices (£mn, Chained Volume measure, 2013 prices)
5. **pop**:  Population (000s)
6. **hours**: Average weekly hours worked (hours)
7. **unemploy**: Unemployment rate (%)
8. **interest**: Bank of England Interest Rate (% calendar year average)
9. **house**: House price index (Calendar year average, Jan 2015 = 100, NB uses building costs as a rough proxy prior to 1895)
10. **exports**: Export Prices, Spliced export price deflator
11. **imports**: Import Prices, Spliced import price deflator
12. **oil**: oil price $ per barrel
13. **CPI**: Headline Consumer Price Index - alternative measure that maximises use of consumer price indices and minimises use of implied consumption deflator.
14. **inflation**: % increase in CPI since previous year
15. **assets**: (£10mn) - total Bank of England assets - indicator of quantitative easing.
16. **balance_sheet++** - assets as % of GDP
17. **deficit_perc** - deficit as a % of GDP


Structure?

```{r str, echo=FALSE}
str(bank)
```

Descriptive Stats?

```{r desc, echo=FALSE}
summary(bank)
```

#2. Further Analysis - some simple plots to looks for patterns and correlations

a) Plot for each variable by year:

```{r plots, echo=FALSE,message=FALSE,warning=FALSE}

bank$assets <- bank$assets/1000000
library(ggplot2)
library(gridExtra)

plotall <- function(dataframe,col) {
g1 <- ggplot(aes(x=year,y=deficit),data = dataframe) + 
  geom_line(color=col)
g2 <- ggplot(aes(x=year,y=surplus),data = dataframe) + 
  geom_line(color=col)
g3 <- ggplot(aes(x=year,y=GDP),data = dataframe) + 
  geom_line(color=col)
g4 <- ggplot(aes(x=year,y=pop),data = dataframe) + 
  geom_line(color=col)
g5 <- ggplot(aes(x=year,y=hours),data = dataframe) + 
  geom_line(color=col)
g6 <- ggplot(aes(x=year,y=unemploy),data = dataframe) + 
  geom_line(color=col)
g7 <- ggplot(aes(x=year,y=interest),data = dataframe) + 
  geom_line(color=col)
g8 <- ggplot(aes(x=year,y=house),data = dataframe) + 
  geom_line(color=col)
g9 <- ggplot(aes(x=year,y=oil),data = dataframe) + 
  geom_line(color=col)
g10 <- ggplot(aes(x=year,y=CPI),data = dataframe) + 
  geom_line(color=col)
g11 <- ggplot(aes(x=year,y=inflation),data = dataframe) + 
  geom_line(color=col)
g12 <- ggplot(aes(x=year,y=exports),data = dataframe) + 
  geom_line(color=col)
g13 <- ggplot(aes(x=year,y=imports),data = dataframe) + 
  geom_line(color=col)
g14 <- ggplot(aes(x=year,y=assets),data = dataframe) + 
  geom_line(color=col)

grid.arrange(g1,g2,g3,g4,g5,g6,g7,g8,g9,g10,g11,g12,g13,g14)
}

plotnew <- function(dataframe,col) {
g1 <- ggplot(aes(x=year,y=deficit),data = dataframe) + 
  geom_line(color=col)
g2 <- ggplot(aes(x=year,y=GDP),data = dataframe) + 
  geom_line(color=col)
g3 <- ggplot(aes(x=year,y=pop),data = dataframe) + 
  geom_line(color=col)
g4 <- ggplot(aes(x=year,y=unemploy),data = dataframe) + 
  geom_line(color=col)
g5 <- ggplot(aes(x=year,y=interest),data = dataframe) + 
  geom_line(color=col)
g6 <- ggplot(aes(x=year,y=house),data = dataframe) + 
  geom_line(color=col)
g7 <- ggplot(aes(x=year,y=oil),data = dataframe) + 
  geom_line(color=col)
g8 <- ggplot(aes(x=year,y=inflation),data = dataframe) + 
  geom_line(color=col)
g9 <- ggplot(aes(x=year,y=assets),data = dataframe) + 
  geom_line(color=col)

grid.arrange(g1,g2,g3,g4,g5,g6,g7,g8,g9)
}

plothist <- function(dataframe,col) {
g1 <- qplot(data = dataframe, bins=20, x = deficit,xlab = 'deficit') 
g2 <- qplot(data = dataframe, bins=20, x = surplus,xlab = 'surplus') 
g3 <- qplot(data = dataframe, bins=20, x = GDP,xlab = 'GDP') 

g4 <- qplot(data = dataframe, bins=20, x = pop,xlab = 'Population')
g5 <- qplot(data = dataframe, bins=20, x = hours,xlab = 'Hours Worked')
g6 <- qplot(data = dataframe, bins=20, x = unemploy,xlab = 'Unemployment')
g7 <- qplot(data = dataframe, bins=20, x = interest,xlab = 'Interest rate')
g8 <- qplot(data = dataframe, bins=20, x = house,xlab = 'House Price Index')
g9 <- qplot(data = dataframe, bins=20, x = oil,xlab = 'Oil Price')
g10 <- qplot(data = dataframe, bins=20, x = CPI,xlab = 'CPI')
g11 <- qplot(data = dataframe, bins=20, x = inflation,xlab = 'Inflation')
g12 <- qplot(data = dataframe, bins=20, x = exports,xlab = 'exports')
g13 <- qplot(data = dataframe, bins=20, x = imports,xlab = 'Imports')
g14 <- qplot(data = dataframe, bins=20, x = assets,xlab = 'Assets')

grid.arrange(g1,g2,g3,g4,g5,g6,g7,g8,g9,g10,g11,g12,g13)
}

plotall(bank,'black')
```

b) Looks like most of the action is after 1900?

```{r 1900plots, echo=FALSE,message=FALSE,warning=FALSE}
bank1900 <-  subset(bank , year > 1900)
plotall(bank1900,'green')
```

c) How about the last 10 years?

```{r 1995plots, echo=FALSE,message=FALSE,warning=FALSE}
bank1900 <-  subset(bank , year > 1995)
plotall(bank1900,'blue')
```

d) And lastly, histograms:

```{r histplots, echo=FALSE,message=FALSE,warning=FALSE}
plothist(bank)
```

Observations at this stage:

* For all variables except inflation, levels seem to be constant, or steadily rising (population, GDP) up to 1900.

After 1900 data shows patterns emerging between:

* deficit and surplus

* GDP, population, CPI, House prices

* unemployment, exports, imports and oil follow a similar pattern but have dipped in the last couple of years

* interest and inflation

* hours worked is not comparable with the other variables

#3. What is my data story

I decided it would be interesting to see how key data fields behaved in years where major economic events had occured.  

I found a list of major economic events on Wikipedia ([see chart at bottom of this page](https://en.wikipedia.org/wiki/Stock_market_crash))

Then I cut down my variables removing the following for the time being:

* Surplus - as it's strongly linked to deficit
* Hours, Imports, Exports, Oil, House Price, CPI,Population, Unemployment - not directly linked, dependent on other factors

I then plotted a few graphs looking at my new paired down data and a few events, multiplying interest and inflation and dividing GDP for comparison.

```{r crashplots, echo=FALSE,message=FALSE,warning=FALSE}
newbank <- bank[c("deficit","year", "assets", "GDP","interest","inflation")] 

write.csv(newbank, file = "new_bank.csv")
 
 ggplot(newbank, aes(x = year)) + 
  geom_line(aes(y = (deficit-(deficit*2)),colour='deficit(+)')) + scale_x_continuous(limits = c(1700, 2015)) +
  geom_line(aes(y = GDP/12,colour='GDP/12')) + 
    geom_line(aes(y = interest*3000,colour='interest*3000')) + 
    geom_line(aes(y = inflation*500,colour='inflation*500')) +
   geom_line(aes(y = assets/4,colour='assets/4')) +
  ylab(label="deficit") + 
  xlab("All of Them")
 

ggplot(newbank, aes(x = year)) + 
  geom_line(aes(y = (deficit-(deficit*2)),colour='deficit(+)')) + scale_x_continuous(limits = c(1700, 1800),breaks = c(1700,1720,1750,1792)) +
  geom_line(aes(y = GDP/12,colour='GDP/12')) + 
    geom_line(aes(y = interest*3000,colour='interest*3000')) + 
    geom_line(aes(y = inflation*500,colour='inflation*500')) +
   geom_line(aes(y = assets,colour='assets')) +
  ylab(label="deficit") + 
  xlab("1720 South Sea Bubble, Panic of 1792")


ggplot(newbank, aes(x = year)) + 
  geom_line(aes(y = (deficit-(deficit*2)),colour='deficit(+)')) +
  scale_x_continuous(limits = c(1800, 1850),breaks = c(1800,1825,1837,1847)) + 
  geom_line(aes(y = GDP/12,colour='GDP/12')) + 
    geom_line(aes(y = interest*3000,colour='interest*3000')) + 
    geom_line(aes(y = inflation*500,colour='inflation*500')) + 
  ylab(label="deficit") + 
  xlab("Panic of 1819, 1825, 1837 and 1847")

ggplot(newbank, aes(x = year)) + 
  geom_line(aes(y = (deficit-(deficit*2)),colour='deficit(+)')) +
  scale_x_continuous(limits = c(1990, 2015),breaks = c(1800,1825,1837,1847)) + 
  geom_line(aes(y = GDP/12,colour='GDP/12')) + 
    geom_line(aes(y = interest*3000,colour='interest*3000')) + 
    geom_line(aes(y = inflation*500,colour='inflation*500')) + 
  ylab(label="deficit") + 
  xlab("1990-2015")



```


#4. The Data Story

It looks like inflation, interest and GDP are  most responsive to economic events.

*My story* From 1720 to date, how has inflation, interest and GDP behaved before and after major economic events.  

New table dimensions:

1.  Crash Year
2.  Description
3.  Average Inflation 5 years before
4.  Average Inflation 5 years after
5.  Average Interest 5 years before
6.  Average Interest 5 years after
7.  Average GDP 5 years before
8.  Average GDP 5 years after 
9.  Deficit as a % of GDP
10.  Assets as a % of GDP

At this stage, I started a python file to create the average value fields and scale my variables from lowest(0) to highest(100) so that I could display them on the same x/y axes.  

As the main purpose of this exercise was to embed D3.js skills I decided to have a quick look at correlations and stop the analysis here. 

Maybe something to come back to at a later date?  

```{r newbank_withcrashes, echo=FALSE,message=FALSE,warning=FALSE}
setwd('/Users/bryonymiles/documents/dataprojects/eda')
new_crashes <-read.csv('bank_final.csv')


summary(new_crashes)


ggplot(new_crashes, aes(x = year)) + 
  scale_x_continuous(limits = c(1700, 2015)) +
    geom_line(aes(y = int_before,colour='int_before')) + 
      geom_line(aes(y = int_after,colour='int_after')) + 
  ylab(label="interest rate") + 
  xlab("Just Interest")
 
ggplot(new_crashes, aes(x = year)) + 
  scale_x_continuous(limits = c(1700, 2015)) +
    geom_line(aes(y = inf_before,colour='inf_before')) + 
      geom_line(aes(y = inf_after,colour='inf_after')) + 
  ylab(label="inflation rate") + 
  xlab("Just Inflation")

ggplot(new_crashes, aes(x = year)) + 
  scale_x_continuous(limits = c(1700, 2015)) +
    geom_line(aes(y = GDP_before,colour='GDP_before')) + 
      geom_line(aes(y = GDP_after,colour='GDP_after')) + 
  ylab(label="£ mn") + 
  xlab("Just GDP")

ggplot(new_crashes, aes(x = year)) + 
  scale_x_continuous(limits = c(1700, 2015)) +
    geom_line(aes(y = inf_before_scale,colour='inf_before scaled')) + 
      geom_line(aes(y = inf_after_scale,colour='inf_after scaled')) + 
    geom_line(aes(y = int_before_scale,colour='int_before scale')) + 
      geom_line(aes(y = int_after_scale,colour='int_after scaled')) + 
    geom_line(aes(y = GDP_before_scale,colour='GDP_before scaled')) + 
      geom_line(aes(y = GDP_after_scale,colour='GDP_after scaled')) + 
  ylab(label="lowest to highest") + 
  xlab("Inflation, Interest and GDP change scaled lowest(0) to highest(100)")
```
